A Mechanised Library for Generic (Co)Programming

### 2024-03-12 paper/paper.tex
typo
Examples

\newcommand{\CatC}{\mathcal{C}}
\newcommand{\CatD}{\mathcal{D}}
\newcommand{\CatE}{\mathcal{E}}
\newcommand{\CatI}{\mathcal{I}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\iso}{\cong}
\newcommand{\ceiling}[1]{\lceil #1 \rceil}
\newcommand{\floor}[1]{\lfloor #1 \rfloor}
\newcommand{\pair}[2]{\langle #1, #2 \rangle}
\newcommand{\haskell}[1]{\mintinline{haskell}{#1}}
%\newcommand{\coq}[1]{\mintinline{coq}{#1}}
\newmintinline[coq]{coq}{fontsize=\small}
\newmintinline[ocaml]{OCaml}{fontsize=\small}
\newminted{coq}{fontsize=\small}

@@

  The key contribution of this paper is an axiom-free framework for generic
programming with recursion schemes in Coq, focusing on proving algebraic laws,
reasoning about program equivalences, and extracting safe, idiomatic code. With
this framework, we formalize a series of divide-and conquer, dynamic
programming, and mutual recursion programs as hylomorphisms. We demonstrate that
the extracted OCaml code for the programs formalized in our framework is
efficient, resembles code that a human programmer might write, and contains no
uses of \ocaml{Obj.magic}. We also present a machine-checked proof of the
well-known short-cut fusion optimsation.

@@

% Similarly to the rest of the definitions, we prove that recursive anamorphisms
% are respectful, and use this fact to build them as higher order proper
% morphisms, and prove their universal law, i.e.\ that recursive anamorphisms are
% the unique solution to the equation
% \coq{f =e l_in \o fmap f \o h},
% where \coq{h} is a recursive coalgebra.
\section{Extraction}
This section covers how to use our approach to calculate recursive programs
from simple, non-recursive specifications. The steps are the following:
\begin{enumerate}
  \item Write a specification as a morphism \coq{f : A ~> B}.
  \item Use Coq's generalised rewriting and the laws of morphisms to 
    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
  \item Project the function \coq{h : A -> B} from \coq{g}
    (i.e. \coq{h = app g}).
  \item Extract \coq{h} to OCaml.
\end{enumerate}
Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
proofs from the code, and inline our definitions. To ease the application of
these steps, we provide the following helper type:
\begin{coqcode}
  Record Ext (f : A ~> B) := 
    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
\end{coqcode}
Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
will be the function \coq{target} that is extensionally equal to \coq{f}.

\subsection{Sorting Algorithms}
\dcas{Talk About this}
\begin{minted}{OCaml}
  let rec msort x = match x with
| [] -> x | _ :: l -> (match l with | [] -> x | _ :: _ ->
     let (l0, r) = splitL x [] [] in
     let cx = fun p -> match p with | Lbranch -> l0 | Rbranch -> r in
     mergeL (msort (cx Lbranch)) (msort (cx Rbranch)))
\end{minted}

@@

\subsection{Harper's Regular-Expression Matcher}

\dcas{foldr/build examples from original paper (or hylo papers)}

>>>>

\newcommand{\CatC}{\mathcal{C}}
\newcommand{\CatD}{\mathcal{D}}
\newcommand{\CatE}{\mathcal{E}}
\newcommand{\CatI}{\mathcal{I}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\iso}{\cong}
\newcommand{\ceiling}[1]{\lceil #1 \rceil}
\newcommand{\floor}[1]{\lfloor #1 \rfloor}
\newcommand{\pair}[2]{\langle #1, #2 \rangle}
\newcommand{\haskell}[1]{\mintinline{haskell}{#1}}
%\newcommand{\coq}[1]{\mintinline{coq}{#1}}
\newmintinline[coq]{coq}{fontsize=\small}
\newmintinline[ocaml]{OCaml}{fontsize=\small}
\newminted{coq}{fontsize=\small}
\newminted{ocaml}{fontsize=\small}

@@

The key contribution of this paper is an axiom-free framework for generic
programming with recursion schemes in Coq, focusing on proving algebraic laws,
reasoning about program equivalences, and extracting safe, idiomatic code. With
this framework, we formalize a series of divide-and conquer, dynamic
programming, and mutual recursion programs as hylomorphisms. We demonstrate that
the extracted OCaml code for the programs formalized in our framework is
efficient, resembles code that a human programmer might write, and contains no
uses of \ocaml{Obj.magic}. We also present a machine-checked proof of the
well-known short-cut fusion optimisation.

@@

% Similarly to the rest of the definitions, we prove that recursive anamorphisms
% are respectful, and use this fact to build them as higher order proper
% morphisms, and prove their universal law, i.e.\ that recursive anamorphisms are
% the unique solution to the equation
% \coq{f =e l_in \o fmap f \o h},
% where \coq{h} is a recursive coalgebra.
\section{Extraction}
This section covers how to use our approach to calculate recursive programs
from simple, non-recursive specifications. The steps are the following:
\begin{enumerate}
  \item Write a specification as a morphism \coq{f : A ~> B}.
  \item Use Coq's generalised rewriting and the laws of morphisms to 
    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
  \item Project the function \coq{h : A -> B} from \coq{g}
    (i.e. \coq{h = app g}).
  \item Extract \coq{h} to OCaml.
\end{enumerate}
Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
proofs from the code, and inline our definitions. To ease the application of
these steps, we provide the following helper type:
\begin{coqcode}
  Record Ext (f : A ~> B) := 
    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
\end{coqcode}
Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
will be the function \coq{target} that is extensionally equal to \coq{f}.  We
use a shortcut tactic, \coq{calculate}, that is a synonym to 
\coq{eapply MkExt}. As a result of this tactic, we will be given the goal 
\coq{f =e ?Goal0}. We can rewrite \coq{f}, unfold definitions, etc, and
complete the definition with \coq{reflexivity}, and this will instantiate the
metavariable \coq{?Goal0} to our desired function. We will use this extensively
in our examples.

\subsubsection{Example 1}
Shortcut deforestation can be expressed succintly in terms of hylomorphisms and
their laws~\cite{TakanoM95}. In particular, given a function:
\begin{coqcode}
  s  : forall A. (App F A -> A) -> (App F A -> A)
\end{coqcode}
We can conclude, by parametricity, that
\begin{coqcode}
  hylo a l_out \o hylo (sigma l_in) c =e hylo (s a) c
\end{coqcode}
This is generally known as the \emph{acid rain} theorem. Unfortunately, this is
\textbf{not} provable in Coq if we want to remain axiom-free, since we would
need to add the necessary parametricity
axiom~\cite{keller_et_al:LIPIcs.CSL.2012.381}.

However, we prove a specific version of this theorem for the list generating
container (i.e.\ the container whose least fixed point is a list), and use it
to encode the example from Takano and Meijer~\cite{TakanoM95}:
\begin{coqcode}
Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
\end{coqcode}
Here, we are defining function \coq{sf1} as the composition of \coq{length},
\coq{Lmap f} and \coq{append ys}. Functions \coq{length} and \coq{Lmap} are
catamorphisms. Function \coq{append ys} is also a catamorphism that
appends \coq{ys} to an input list. It is defined by applying an algebra to
every cons node of a list, and applying a catamorphism with the input algebra
to \coq{ys} in the nil case:
\begin{coqcode}
Definition tau (l : list A) (a : Alg (ListF A) B) : App (ListF A) B -> B :=
  fun x => match x with | MkCont sx kx => match sx with
  | s_nil => fun _ => (hylo a ilist_coalg) l
  | s_cons h => fun kx => a (MkCont (s_cons h) kx)
  end kx end.
Definition append (l : list A) := hylo (tau l l_in) ilist_coalg.
\end{coqcode}
Here, \coq{ilist_coalg} is a recursive coalgebra from Coq lists to the
\coq{ListF} container. We apply the hylo fusion law repeatedly, unfold
definitions, and simplify in our specification for \coq{sf1}: 
\begin{coqcode}
Definition sf1 (f : A ~> B) ys : Ext (length \o Lmap f \o append ys).
  rewrite hylo_map_fusion, <- acid_rain.
  simpl; reflexivity.
Defined.
\end{coqcode}
From this, we extract the following OCaml code:
\begin{ocamlcode}
let rec sf1 f ys = function | [] -> let rec f0 = function
                                      | [] -> (Uint63.of_int (0))
                                      | _ :: t -> add (Uint63.of_int (1)) (f0 t)
                                    in f0 ys
                            | _ :: t -> add (Uint63.of_int (1)) (sf1 f ys t)
\end{ocamlcode}

\subsubsection{Example 2} We then prove that the length function fuses with
the naive quadratic \emph{reverse} function:
\begin{coqcode}
Definition sf2 : Ext (length \o reverse). 
  calculate.  unfold length, reverse. rewrite hylo_fusion_l. 
  2:{ (* Rewrite into the fused version *) }
  simpl; reflexivity.
Defined.
\end{coqcode}
This code extracts to the optimised length function on the input list:
\begin{ocamlcode}
let rec ex2 = function | [] -> (Uint63.of_int (0))
                       | _ :: t -> add (Uint63.of_int (1)) (ex2 t)
\end{ocamlcode}

\subsection{Sorting Algorithms}

So far, our examples could be modelled exclusively using catamorphisms. Here,
we show examples that require recursive hylomorphisms and termination proofs.

\subsubsection{Mergesort}

\begin{minted}{OCaml}
  let rec msort x = match x with
| [] -> x | _ :: l -> (match l with | [] -> x | _ :: _ ->
     let (l0, r) = splitL x [] [] in
     let cx = fun p -> match p with | Lbranch -> l0 | Rbranch -> r in
     mergeL (msort (cx Lbranch)) (msort (cx Rbranch)))
\end{minted}

\subsubsection{Quicksort}

\subsubsection{Fusing a divide-and-conquer computation}

@@
